---
title: "plotly"
output: 
  html_document:
    keep_md: true
    theme: paper
---


```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

You can make interactive graphs and plots in R using the plotly library. This is a demo of what plotly is and why it’s useful, and then you can try converting a static ggplot graph into an interactive plotly graph.

```{r}
library(tidyverse)
library(gapminder)
library(plotly) 
```

It’s very easy to convert an existing ggplot2 graph into an interactive graph with plotly::ggplotly

On the below graph, explore the interactive options:

* Hover your cursor over individual points
* Zoom in and out by dragging across / using the zoom tool
* Single- and double-click items on the legend to isolate groups of points
* While zoomed-in, use the pan tool to “move” around the plot, google maps style!

```{r}
p <- gapminder %>%
    ggplot(aes(x = gdpPercap, y = lifeExp, color = continent)) +
    geom_point() 

p %>%
    ggplotly()
```

We can also make interactive graphs using the the `plotly::plot_ly` function:

```{r}
p <- gapminder %>%
    plot_ly(x = ~gdpPercap,
            y = ~lifeExp,
            color = ~continent,
            
            # mode specifies the geometric object e.g. "markers" for points, "line" for lines
            mode = 'markers',
            
            # type controls the "type" of graph e.g. 'bar', 'scatter'
            type = 'scatter'
            ) %>% 
  print
```

To share with others:

1 Create a plotly account @ plot.ly
2 Navigate to settings, and take in the following information:
* your user name
* api key

Now, we will tell R your account information so that we can upload these plots to the web.

(Note that once we run api_create(), the browser will open to a webpage displaying your interactive plot. You can share this page with others, but they will only be able to view. If you want others to be able to edit the graph, you can invite others to “collaborate” in the “Sharing link” option.)

```{r}
# fill in the below with your information
Sys.setenv("plotly_username"="your_plotly_username")
Sys.setenv("plotly_api_key"="your_api_key")

# upload our plots to the website
api_create(p, filename = 'name-of-your-plot')
```

