---
title: "cm005 Worksheet: Exploring Geometric Objects"
output: 
  html_document:
    keep_md: true
    theme: paper
---

## Preliminary

Begin by loading the required packages. If you don't have these installed (or don't know whether you have them installed), you can install them by executing the following code in your console. This code will not run (it's not marked as "r"), but this format is easier to copy and paste than the commented-out format.

```
install.packages("tidyverse")
install.packages("scales")
install.packages("tsibble")
```

Now run this code chunk to load the packages:

```{r load packages, warning = FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(tsibble))
knitr::opts_chunk$set(fig.align = "center")
```

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## Exercise 0: Scatterplot from class guide

This plot was run line-by-line to see how the different layers of ggplot get added on top of each other (section 5.6).  Spoilers, the "+" sign is important.  Also, "alpha" refers to the transparency of overlapping points.

``` {r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  theme_bw() +
  ylab("Life Expectancy")

#alternative for colorcoding by another grouping variable:
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point(mapping = aes(color=continent),alpha = 0.1)
#note that the graph's color-to-continent key is still done with points at that same transparency, so they're hard to see.  You can probably change this manually?

```
Double colons are the syntax for "pull this function from this specific package", because multiple packages sometimes have the same name for different functions.

See the guidebook for the table describing the different pieces of grammar in this plot (section 5.5)

## Exercise 1: Bar Chart Grammar (Together)

Consider the following plot. Don't concern yourself with the code at this point.

```{r, fig.width = 5, fig.height = 2}
gapminder %>% 
  filter(year == 2007) %>% 
  mutate(continent = fct_infreq(continent)) %>% 
  ggplot(aes(continent)) +
  geom_bar() +
  theme_bw()
```

Fill in the seven grammar components for this plot.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder`, only data from 2007 |
| __aesthetic mapping__ | x = `continent`, y = `..count..` |
| __geometric object__  | bar |
| scale                 | linear |
| statistical transform | none |
| coordinate system     | rectangular |
| facetting             | none |

## Exercise 2: `ggplot2` Syntax (Your Turn)

The following is a tsibble (a special type of tibble containing time series data, which we'll see more of later), stored in the variable `mauna`, of CO$_2$ concentrations collected monthly at the Mauna Loa station.

Execute this code to store the data in `mauna`:

```{r}
(mauna <- tsibble::as_tsibble(co2) %>% 
   rename(month = index, conc = value))
```

### 2(a)

Produce a line chart showing the concentration over time. Specifically, the plot should have the following grammar components:

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `mauna` |
| __aesthetic mapping__ | x: month, y: conc |
| __geometric object__  | lines |
| scale                 | linear |
| statistical transform | none |
| coordinate system     | rectangular |
| facetting             | none |

Fill in the blanks to obtain the plot:

```{r, fig.width = 5, fig.height = 2}
ggplot(mauna, aes(month, conc)) +
  geom_line() +
  ylab("CO2 concentration") +
  xlab("Year") +
  theme_minimal()
```

### 2(b)

It turns out that you're allowed to specify the aesthetic mappings in a `geom` layer instead of, or in addition to, in the `ggplot()` function, with the following rules:

- Aesthetics appearing in a `geom` layer apply only to that layer.
- If there are conflicting aesthetics in both the `ggplot()` function and the `geom` layer, the `geom` layer takes precedence.

The following code mistakenly puts the month variable on the y-axis. Fill in the `FILL_THIS_IN` so that you still obtain the same result as above.

```{r, fig.width = 5, fig.height = 2}
ggplot(mauna, aes(y = month)) +
  geom_line(aes(month, conc)) +
  ylab("conc")
```

### 2(c)

You can store the output of the plot in a variable, too. Store the plot from 2(a) in the variable named `p`, then add a layer to `p` that adds green points to the plot.

```{r, fig.width = 5, fig.height = 2}
p <- ggplot(mauna, aes(month, conc)) +
  geom_line() +
  ylab("CO2 concentration") +
  xlab("Year") +
  theme_minimal()
p +
  geom_point(colour = "green")
```

### 2(d)

What's wrong with the following code? Fix it. (answer, it's missing the aes arguement.  Interestingly, that does NOT have go in the first ggplot call.)

```{r, fig.width = 5, fig.height = 2}
ggplot(gapminder) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp), alpha = 0.1)
```


### 2(e) BONUS

So you're a ggplot2 pro? Then, let's see this plot adapted to polar coordinates. Specifically:

- angle is month (January through December)
- radius is CO$_2$ concentration

The plot should look like a spiral, or concentric circles. 

```{r, fig.width = 5, fig.height = 2}
p +
  coord_polar()
```

